Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2025-01-31 paper.qmd
fig 5 re-done

#preferred solution 5 clusters using dunn/dunn2 index
set.seed(150)
X_km <- kmeans(X_outside_scaled, 5)
# making a vector that labels all points
# use 0 to indicate point is not an outlier
cluster_labels_all <- as_tibble(X,
                               rownames = "year") |>
  left_join(as_tibble(X_km$cluster,
                              rownames = "year")) |>
  mutate(cluster = if_else(is.na(value), 0, value)) |>
  arrange(cluster)
cluster_labels <- factor(cluster_labels_all$cluster,
                         levels = c(1:5, 0))

@@

set.seed(22)
animate_xy(as.matrix(X_sorted), 
           guided_anomaly_tour(anomaly_index(),
                               ellipse=as.matrix(Sigma),
                               ellc = size_weather,
                               ellmu = mu), 
           ellipse=as.matrix(Sigma), ellc = size_weather,
           ellmu = mu, cex=2, 
           axes="bottomleft",
           col = as.factor(cluster_labels),
           palette = "Batlow",
           center = FALSE, rescale = FALSE,
           ellmarks = FALSE, axislablong = TRUE)

# color codes
#"#201158" "#005260" "#007C46" "#999600" "#F9A391" "#FFCEF4"
# "#005260"
# wait for getting random starting planes for
# better results
set.seed(55)
animate_xy(X_sorted[cluster_labels == 4,], 
           guided_anomaly_tour(anomaly_index(),
                               ellipse=as.matrix(Sigma),
                               ellc = size_weather,
                               ellmu = mu), 
           ellipse=as.matrix(Sigma), ellc = size_weather,
           ellmu = mu,
           axes="bottomleft",
           col = "#999600", half_range = 7, cex = 2,
           center = FALSE, rescale = FALSE,
           ellmarks = FALSE, axislablong = TRUE)

@@

![Only cluster 4](images/weather2.png){#fig-weather2 width=275 align="center"}

Examining weather anomalies using projections identified by the anomaly index: (a) computed on all outlying points, (b) computed only for cluster 4. In (a) the outlying points are somewhat orthogonal to the normal ellipse, which is primarily using the temperature min (`atmn`) and max (`atmx`) variables. In (b) the two observations are primarily extreme in the minimum temperature (`atmn`) variable, and have lower than normal minima.
:::

>>>>

#preferred solution 5 clusters using dunn/dunn2 index
set.seed(150)
X_km <- kmeans(X_outside_scaled, 5)
# making a vector that labels all points
# use 0 to indicate point is not an outlier
cluster_labels_all <- as_tibble(X,
                               rownames = "year") |>
  left_join(as_tibble(X_km$cluster,
                              rownames = "year")) |>
  mutate(cluster = if_else(is.na(value), 0, value)) |>
  arrange(cluster)
cluster_labels <- factor(cluster_labels_all$cluster,
                         levels = c(1:5, 0))
ptshape <- factor(ifelse(cluster_labels_all$cluster>0, "out", "in"))

@@

set.seed(22)
animate_xy(as.matrix(X_sorted), 
           guided_anomaly_tour(anomaly_index(),
                               ellipse=as.matrix(Sigma),
                               ellc = size_weather,
                               ellmu = mu), 
           ellipse=as.matrix(Sigma), ellc = size_weather,
           ellmu = mu, cex=2, 
           axes="bottomleft",
           col = as.factor(cluster_labels),
           pch = ptshape,
           palette = c("#040404", "#55125D", "#A01975", "#E45358", "#F7AA2D", "#559999"),
           shapeset = c(1, 16),
           center = FALSE, rescale = FALSE,
           ellmarks = FALSE, axislablong = TRUE)

# color codes
#"#201158" "#005260" "#007C46" "#999600" "#F9A391" "#FFCEF4"
# "#005260"
# wait for getting random starting planes for
# better results
set.seed(55)
animate_xy(X_sorted[cluster_labels == 4,], 
           guided_anomaly_tour(anomaly_index(),
                               ellipse=as.matrix(Sigma),
                               ellc = size_weather,
                               ellmu = mu), 
           ellipse=as.matrix(Sigma), ellc = size_weather,
           ellmu = mu,
           axes="bottomleft",
           col = "#E45358", pch=16, half_range = 7, cex = 2,
           center = FALSE, rescale = FALSE,
           ellmarks = FALSE, axislablong = TRUE)

@@

![Focus on cluster 4](images/weather2.png){#fig-weather2 width=275 align="center"}

Examining weather anomalies using the anomaly index guided tour to assess what combinations of variables are creating the extremes. Anomalies were grouped based on similarity in direction, and colour indicates cluster membership.  In (a) the anomaly index was computed on all outlying points, and most clusters (except for cluster 4) are outlying points are somewhat orthogonal to the normal ellipse. This extremeness is primarily due to temperature min (`atmn`) and max (`atmx`) variables as read from the axes. In (b) only the two points in cluster 4 are used for computing the index. These two observations are primarily extreme in the minimum temperature (`atmn`) variable, and have lower than normal minima.
:::

### 2025-01-31 paper.tex
fig 5 re-done

\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/weather1.png}

@@

\includegraphics[width=3.125in,height=\textheight,keepaspectratio]{images/weather2.png}

@@

\subcaption{\label{fig-weather2}Only cluster 4}

@@

\caption{\label{fig-weather}Examining weather anomalies using
projections identified by the anomaly index: (a) computed on all
outlying points, (b) computed only for cluster 4. In (a) the outlying
points are somewhat orthogonal to the normal ellipse, which is primarily
using the temperature min (\texttt{atmn}) and max (\texttt{atmx})
variables. In (b) the two observations are primarily extreme in the
minimum temperature (\texttt{atmn}) variable, and have lower than normal
minima.}

>>>>

\includegraphics[width=2.86458in,height=\textheight,keepaspectratio]{images/weather1.png}

@@

\includegraphics[width=2.86458in,height=\textheight,keepaspectratio]{images/weather2.png}

@@

\subcaption{\label{fig-weather2}Focus on cluster 4}

@@

\caption{\label{fig-weather}Examining weather anomalies using the
anomaly index guided tour to assess what combinations of variables are
creating the extremes. Anomalies were grouped based on similarity in
direction, and colour indicates cluster membership. In (a) the anomaly
index was computed on all outlying points, and most clusters (except for
cluster 4) are outlying points are somewhat orthogonal to the normal
ellipse. This extremeness is primarily due to temperature min
(\texttt{atmn}) and max (\texttt{atmx}) variables as read from the axes.
In (b) only the two points in cluster 4 are used for computing the
index. These two observations are primarily extreme in the minimum
temperature (\texttt{atmn}) variable, and have lower than normal
minima.}
